---
title: "No Correlation Between Gun Control Legislation and Amount of Shootings in Toronto, Ontario"
author: "Russell Luchin"
thanks: "Code, data, and .QMD available at: https://github.com/rluchin/No-Correlation-Between-Gun-Control-and-Shootings. All code and analysis done with R programming language (R Core Team, 2023). Credit to Luca Carnegie for the code template for this footer."
date: 01-26-2024
date-format: long
abstract: "Data from the Toronto Police Service was used to analyze year-to-year trends on shootings within the city alongside the passage of gun access legislation. Based on the observed trends, there is no significant correlation between passing or reversing gun control legislation on the number of shootings in Toronto. Analysis of Toronto corroberates similar studies based on gun access legislation in the United States, indicating a broader pattern of gun access legislation being unrelated to total shooting deaths/injuries."
bibliography: bibliography.bib
format: pdf
editor: visual
---

## Table of Contents

**1 Introduction**

**2 Data**

2.1 Raw Data from Toronto Police Service\
2.2 Cleaned Data\
2.3 Methodology

**3 Results**

3.1 Shootings by Years With Major Gun Control Passing\
3.2 Total Shootings, Year-Over-Year, 2007-2022\
3.3 Shootings By Month, 2019\
3.4 Shootings per Year, 2008 to 2011

**4 Conclusion**

# Introduction

Gun violence is an epidemic affecting the vast majority of North American cities [@citeNorth]. The most common piece of legislation we see debated in government is that of firearms access; this debate has become incredibly divisive and leads to a sharp divide in policy-making between party lines. However, there is a debatable degree of impact that this type of legislation has, as a whole, on total shooting trends in major cities.

This paper visualizes and observes total shooting trends in Toronto alongside the passage of major firearms access policies, both restrictive and non-restrictive policies, using the R programming language [@citeR]. The purpose of doing this is to observe whether firearms access legislation has a direct impact on shooting trends within Toronto, a city which has observed a steady increase in shootings in the face of varying gun-control policies being written into law.

Findings indicate that there is limited, if any, correlation between firearms access legislation and overall shooting trends. There is no data to indicate that increasing access to firearms will lead to more shootings, and the restricting access leads to less. As this is the common divide in politics on the issue, it is important to observe this data to promote informed policy making in government. Public safety is an essential issue for any government, but policy that disproportionatly affects indigenous communities [@citeAFN] must be studied for effectiveness, rather than passed quickly for a quick win.

```{r}
#| echo: false
#| include: false
## Preamble
# Purpose: Find trends within number of shootings in Toronto
# Author: Russell Luchin
# Email: russell.luchin@mail.utoronto.ca
# Date: 24 January, 2024

## Workspace Setup
library(knitr)
library(janitor)
library(lubridate)
library(opendatatoronto)
library(tidyverse)
library(dplyr)
```

# Data

## **2.1 Raw Data**

Raw data was acquired through Open Data Toronto [@citeToronto]. The specific data set acquired is the official Toronto Police Service daily shootings tracker. This data set is extensive, cataloging every firearm discharge within city limits alongside Case ID and amount of injuries and/or deaths associated with the event. One thing of note is, prior to 2008, the tracker only contains the data for illegal firearm discharges within city limits, as Toronto didn't pass the bylaw against discharging firearms until 2008 [@cite2008]. Furthermore, the uncleaned dataset includes specific elements of each individual shooting; such as hour the shooting took place and within what ward. These variables are irrelevant to us as we are focused on the greater year-to-year trends of shootings within the city. What we focus on are columns 4 and 5, or "OCC_YEAR" and "OCC_MONTH" respectively.

```{r}
#| echo: false
#| include: false
###ACQUIRE

package <- show_package("4bc5511d-0ecf-487a-9214-7b2359ad8f61")
package

toronto_shootings <-
  list_package_resources("4bc5511d-0ecf-487a-9214-7b2359ad8f61") |>
  filter(name ==
           "shootings-firearm-discharges - 4326.csv") |>
  get_resource()

write_csv(
  x = toronto_shootings,
  file = "toronto_shootings.csv"
)

#Review both head and tail of csv file to confirm correct start and end dates
head(toronto_shootings)
tail(toronto_shootings)
```

## **2.2 Cleaned Data**

Data cleaning included eliminating all irrelevant columns folding the relevant columns into a single variable. The cleaned data set ignores discharges with no injuries or deaths, and folds the "Deaths" and "Injuries" variables into a single one labelled "Shootings". Furthermore, the daily shootings were summarized to be organized by total per month of the year, rather than by day of week. The new numbers were checked to be accurate by comparing the generated variables with the official Toronto Police numbers for daily shooting injuries and deaths [@citeTPS].

```{r}
#| echo: false
#| include: false
##DATA CLEANING

# Import dataset to temporary dataframe
shootings_data <- read.csv("toronto_shootings.csv")

# Group the data by Year and Month, and summarize the number of shootings
shootings_summary <- shootings_data %>%
  mutate(Month = format(ymd(paste("2022", OCC_MONTH, "1")), "%b")) %>%
  group_by(OCC_YEAR, Month) %>%
  summarize(Shootings = sum(DEATH + INJURIES), .groups = 'drop')

# Rename columns to Year, Month, and Shootings
toronto_shootings_cleaned <- shootings_summary %>%
  rename(
    Year = OCC_YEAR
  )

# Save the final dataframe to a CSV file
write.csv(toronto_shootings_cleaned, "toronto_shootings_cleaned.csv", row.names = FALSE)

# Ensure the cleaned file has the correct Year/Month/Shootings structure
head(toronto_shootings_cleaned)
tail(toronto_shootings_cleaned)

##CLEANED DATA WAS DOUBLE CHECKED FOR ACCURACY BY CROSS-REFERENCING CLEANED SHOOTINGS WITH THE OFFICIAL TORONTO POLICE TALLIES OF PERSONS INJURED/KILLED FOUND HERE: https://data.torontopolice.on.ca/pages/shootings
```

```{r}
#| echo: false

# Read data from the CSV file
data <- read.csv("toronto_shootings_cleaned.csv")

cat("Table of First 6 Rows from Cleaned Data")

# Display the first 6 rows as a table
kable(head(data, 6), format = "html")
```

## **2.3 Methodology**

Year-to-year and month-to-month trends are used visualize by graphing shooting data alongside relevant legislation concerning legal firearm access. This is achieved by graphing specific dates (years/months), between 2007 and 2022, wherein significant firearm access ("gun control") legislation was passed. This range was chosen to keep the data from getting cluttered while also giving sufficient room to analyze year-over-year trends at any point. Legislation reviewed includes policy which increases legal firearm access ("reducing gun control") and legislation which reduces legal firearm access ("increasing gun control"). For relevant years, total shooting trends will be analyzed by comparing when the legislation was received royal ascent (became law) and what happened in the short-term and long-term following the enactment of said legislation.

# Results

## **3.1** Shootings by Years With Major Gun Control Passing

```{r}
#| echo: false
###EXPLORE

##Create 3 line graphs, graphing total shootings per month, for 2019, 2021, and 2022

# Read the cleaned dataset
shootings_summary <- read.csv("toronto_shootings_cleaned.csv")

# Filter the data for the years 2019, 2021, and 2022
filtered_data <- shootings_summary %>%
  filter(Year %in% c(2019, 2021, 2022))

# Create a line plot with facets for each year
ggplot(data = filtered_data, aes(x = factor(Month, levels = month.abb), y = Shootings, group = Year, color = factor(Year))) +
  geom_line() +
  labs(title = "Shootings by Year, Major Gun Control Passing (Figure 3.1)",
       x = "Month",
       y = "Shootings") +
  scale_color_manual(values = c("2019" = "blue", "2021" = "green", "2022" = "red"),
                     labels = c("2019", "2021", "2022")) +
  theme_minimal() +
  facet_wrap(~Year, ncol = 1) +
  labs(color = "Year")
```

Figure 3.1 shows shootings by month in recent years where major gun control received ascent. This includes 2019, 2021, and 2022; where Bill C-71 [@citeC71], the AR-15 prohibition [@citeOIC], and the handgun transfer freeze [@citeC21], were enacted respectively. What we find is there is no significant change in trends following the enactment of said legislation. Bill C-71 received ascent in June 2019, and we observe a huge spike in shootings following the summer months. However, this specific spike is consistent across all 3 of these graphed years, occurring around the same seasonal change, which rules out direct correlation with C-71 alone. In 2021, the first full year following the AR-15 prohibition in May 2020 we see similar trends to 2019. Notably, shootings for 2021 are lower, however this is also during a period of varying COVID-19 restrictions [@citeCOVID], which makes it unlikely that the decline is due to the prohibition alone. Finally, we observe 2022, where the handgun transfer freeze was enacted in October 2022. We see an immediate decline going into November of that year, but this is consistent with the seasonal trends in both 2021 and 2019. We cannot observe correlation from these years alone.

## 3.2 Total Shootings, Year-Over-Year, 2007-2022

```{r}
#| echo: false
##Create a bar graph for total shootings per year, 2006 to 2023

# Read the cleaned dataset
shootings_summary <- read.csv("toronto_shootings_cleaned.csv")

# Filter the data for the years 2006 to 2023
filtered_data <- shootings_summary %>%
  filter(Year >= 2007 & Year <= 2022)

# Summarize the data to calculate the total shootings per year
total_shootings_per_year <- filtered_data %>%
  group_by(Year) %>%
  summarize(TotalShootings = sum(Shootings))

# Create a bar graph for total shootings per year
ggplot(data = total_shootings_per_year, aes(x = factor(Year), y = TotalShootings)) +
  geom_bar(stat = "identity", fill = "grey") +
  labs(title = "Total Shootings per Year (2007 to 2022) (Figure 3.2)",
       x = "Year",
       y = "Total Shootings") +
  theme_minimal()
```

Figure 3.2 visualizes the full year-over-year shooting trends in the city of Toronto from 2007 to 2022. During this period, 2 conflicting gun control legislative philosophies were tested. We split the analysis between 2 periods: 2007-2015 (during a pro-access government tenure), and 2016-2022 (a restrict-access government tenure)\
\
Working from 2007 onward, we see the first major piece of pro-ownership policy receive ascent in 2012, where the "Long Gun Registry", or a registry of all firearms classified as "Non-Restricted" in the Canadian legal code, was decommissioned and its data destroyed [@citeLGR]. From 2012, we see an unsubstantial decline in shootings in 2013 followed by a drastic decline in 2014. While it is possible that the registry's destruction had a delayed effect, it is unlikely [@citeLGR]. We can extrapolate this data to imply that the long-gun registry, and its destruction, had no real effect on reducing or increasing the amount of shootings in Toronto.\
\
Working from 2016, we see an almost immediate increase in shootings overall in Toronto compared to 2007-2015. While this trend is commonly construed to mean "gun control = gun violence" [@citeCCFR], this trend actually follows a steady increase in shootings from 2014 onward. The first major piece of firearms legislation we see in this period is in 2019, with the ascent of C-71 [@citeC71]. This year is the most violent in recent memory for Toronto, seeing upwards of 250 shootings for that year. Looking at 2019 alone, this could indicate that restricting legal access leads to an increase in gun crime. However, we see a gradual decline in shootings following 2019 and continuing into 2023. While the gradual passing of gun control in these years [@citeC21] could potentially be attributed to the decline, it is dishonest to ignore the circumstances in 2020-2022 where regular COVID restrictions affected public and private gatherings [@citeCOVID].\
\
By looking at nearly 2 decades of shooting trends, we start to see that restricting (or de-restricting) legal firearms access has no definitive correlation with how many shootings are committed year to year.

## 3.3 Shootings By Month, 2019

```{r}
#| echo: false
##Create a line graph for total shootings per month, 2019

# Read the cleaned dataset
shootings_summary <- read.csv("toronto_shootings_cleaned.csv")

# Filter the data for the year 2019
filtered_data_2019 <- shootings_summary %>%
  filter(Year == 2019)

# Create a line graph for shootings per month in 2019
ggplot(data = filtered_data_2019, aes(x = factor(Month, levels = month.abb), y = Shootings, group = 1)) +
  geom_line() +
  labs(title = "Shootings by Month in 2019 (Figure 3.3)",
       x = "Month",
       y = "Shootings") +
  theme_minimal()

##BILL C-75 RECIEVED ROYAL ASCENT JUNE 21st, 2019 (Source: https://www.justice.gc.ca/eng/rp-pr/csj-sjc/jsp-sjp/c75/p3.html)
```

2019 saw a record number of shooting deaths and injuries in the city of Toronto. This spike occured alongide the passage of the most substantial gun control policies at the time [@citeC71]. In Figure 3.3, we see the overall shooting trends remain consistent with what we see year-to-year in the city of Toronto (Figure 3.1). However, there is a more dramatic spike between July and August than is to be expected: a 200%\~ increase as opposed to a 100%\~ rise (Figure 3.1, Figure 3.5). This would indicate a positive correlation between passing gun control and a rise in city-wide shootings, but as this is the only instance of a correlation like in 2 decades of data it's more likely that there is an aggravating factor, unrelated to gun control, which is leading to a rise in shootings. We continue to observe a pattern of firearm access legislation having no correlation with total shootings.

## 3.4 Shootings per Year, 2008 to 2011

```{r}
#| echo: false
##Create a line graph for total shootings per year, 2008 to 2011

# Read the cleaned dataset
shootings_summary <- read.csv("toronto_shootings_cleaned.csv")

# Filter the data for the years 2008 to 2011
filtered_data <- shootings_summary %>%
  filter(Year >= 2008 & Year <= 2011)

# Summarize the data to calculate the total shootings per year
total_shootings_per_year <- filtered_data %>%
  group_by(Year) %>%
  summarize(TotalShootings = sum(Shootings))

# Create a line graph for total shootings per year
ggplot(data = total_shootings_per_year, aes(x = Year, y = TotalShootings)) +
  geom_line() +
  labs(title = "Total Shootings per Year (2008 to 2011) (Figure 3.4)",
       x = "Year",
       y = "Total Shootings") +
  theme_minimal()

```

In 2008, Toronto passed a major piece of municipal legislation which banned the discharge of firearms within city limits [@cite2008]. If we are to look at the year-over-year following 2008 alone, it would be reasonable to assume that there is a direct correlation between the passage of this legislation and a decline in shootings. This would be a concrete instance of gun control reducing total shootings so far in this analysis. However, the year over year doesn't give the full picture.\
\
In Figure 3.5, we see a graph of the monthly shootings in 2008. The legislation was proposed in May, 2008 and was officially adopted as a by-law in June of the same year [@cite2008]. When observing the monthly shootings, we see no change in observed trends, other than a relatively high spike in the summer shootings volume when compared to Figure 3.1. This indicates that the year-to-year decrease from 2008 in shootings was due to factors unrelated to the passage of gun control. We continue to observe limited correlation between firearm access legislation and reduction/increase in shootings.

```{r}
#| echo: false
##Create a line graph for total shootings per month, 2008

# Read the cleaned dataset
shootings_summary <- read.csv("toronto_shootings_cleaned.csv")

# Filter the data for the year 2008
filtered_data_2008 <- shootings_summary %>%
  filter(Year == 2008)

# Summarize the data to calculate the total shootings per month
total_shootings_per_month_2008 <- filtered_data_2008 %>%
  group_by(Month) %>%
  summarize(TotalShootings = sum(Shootings))

# Create a line graph for total shootings per month in 2008
ggplot(data = total_shootings_per_month_2008, aes(x = factor(Month, levels = month.abb), y = TotalShootings, group = 1)) +
  geom_line() +
  labs(title = "Total Shootings per Month in 2008 (Figure 3.5)",
       x = "Month",
       y = "Total Shootings") +
  scale_x_discrete(limits = month.abb) +  # Set the order of months
  theme_minimal()
```

# Conclusion

From the limited sample size of Toronto alone, we observe there to be no definitive correlation between firearms access laws and total shootings in Toronto. There is no correlation to be found between the passage of gun control and the decrease of shootings, and the loosening of gun control and an increase in shootings (or vice-versa). For every potential correlation (Figure 3.4, Figure 3.2), scrutinizing the data leads to said correlation becoming limited at best (Figure 3.5), and unrelated at the worst (Figure 3.2, 2020-2022).

Effective legislation begins at understanding the issue. Toronto continues to see generally low rate of decline in shootings (Figure 3.2), despite regular legislative actions which target firearm access. Policymaking without analyzing the effectiveness (or ineffectiveness) over time has resulted in total shootings being unaffected while law-abiding groups of Canadians become disproportionately affected by policy change [@citeAFN].

Based on the data analyzed, it would be in the government's best interest to understand what causes a spike in gun crime, and how to mitigate it without the disproportionate "punishment" of Canadians who rely on firearms for sustenance or sport [@citeAFN]. This paper corroborates the findings of [@citeRAND] and [@citeDDP], among others who have studied the effects of firearms access legislation and found no correlation with gun violence.\
\
\
Special thanks to the tidyverse [@citeTidy], dplyr [@citeDplyr], lubridate [@citeLube], janitor [@citeJan], knitr [@citeKnitr], and R [@citeR] teams for making this analysis possible.

# References
